Measuring the double-drift illusion with hand tracking: capturing illusion resets

### 2020-08-04 Manuscript.Rmd
adding comparison with Sirui's PSE data

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
# The functions here allow downloading the collected data from the OSF project:
source('R/data.R')
# This file has some common helper functions:
source('R/common.R')
# The functions in here are used on the data of the bounded tracking experiments:
source('R/tracking_bounded.R')
# And finally some code for the re-tracing experiment...
source('R/retrace_onePass_V4.R')
# As well as models fit to that data:
source('R/models.R')
```

@@

@@

Some of our participants seem to perceive "jump resets" though, and the model as published, does not predict any such resets. This could be added by including a causal inference step [e.g. @Wei2011]. For the subset of participants that experience jump resets, there may be a process that compares the two position signals (perceived and retinal) and when these exceed some threshold, decides that the two position signals do not share a common source. At that point the perceived position may be discarded, or reset to the retinal position, and the build-up of the illusion may start again. For the remaining participants, either the discrepancy between retinal and perceived location never exceeds their threshold, or perhaps they do not employ any causal inference process.

@@

We find that online, manual tracking of the double-drift stimulus is sensitive to illusion strength, but doesn't display clear resets. We do find spontaneous resets in delayed, single-pass re-tracing. The illusion appears to reset, or max out, predominantly at a spatial distance from the true stimulus location, suggesting that resets resolve a source attribution conflict between the true (retinal) and illusory (perceived) position of the stimulus.

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
# The functions here allow downloading the collected data from the OSF project:
source('R/data.R')
# This file has some common helper functions:
source('R/common.R')
# The functions in here are used on the data of the bounded tracking experiments:
source('R/tracking_bounded.R')
# And finally some code for the re-tracing experiment...
source('R/retrace_onePass_V4.R')
# As well as models fit to that data:
source('R/models.R')
# And finally, code to compare this data with Sirui's PSEs
source('R/time.R')
```

@@

## Comparison with Sirui's data set

While here we find that a spatial limit is the better explanation for resets, other evidence may point at a temporal limit instead. In an as of yet unpublished study [Sirui] obtained the point of subjective equivalence (PSE) of a double drift stimulus, and a static point that appeared some time after the double drift stimulus disappeared. The double-drift stimulus was on a 2 or 4 DVA length trajectory, and disappeared after 1, 2 or 3 seconds. The PSE's can be modelled very accurately with a model that assumes a fixed chance of a reset per unit of time (~.7%/s), that is: it explains the data using only a temporal bound on the illusion. Interestingly, this fully contradicts the current findings. The model predicts that the PSE decreases over time as it becomes more likely that a reset has occurred. We will compare the two data sets here.

```{r fig8, echo=FALSE, message=FALSE, fig.width=8, fig.height=3.5, fig.cap="\\label{fig:fig8}**Figure 8. PSE's of double-drift position and continuous tracking**"}
plotDeviations()
```

If the PSE's express the average perceived deviations from the actual trajectory of the double-drift stimulus, then they should show a similar pattern as the averaged trajectories in experiment 2. In order to be able to compare the two data sets, we will include all trajectories where our simple algorithm did not detect a reset, which should also be included in the PSE data. Further, we scale the trajectories, so that the Y end-point matches the duration of stimulus. Since the PSE data was collected in degrees visual angle, we convert the unit to degrees visual angles, using the estimate at the centre of the workspace. For the 4-second stimuli, we removed the last second. Figure 8 shows the individual participants' and the averaged trajectories (purple, red, orange) as well as the PSE's obtained in the other study (shown separately for 2 and 4 DVA stimulus trajectories) in gray.

For all internal motion speeds, we see a very similar average trajectory, that does not get closer to veridical over time. We verified this with a repeated measures ANOVA on perceived deviation from veridical stimulus position, using internal motion speed (2, 3, or 4 cycles/s) and time (1, 2, or 3 seconds after stimulus onset) as factors. If the trajectory data from experiment 2 here follows the same pattern as the PSE data, there should be an effect of time, and perhaps an interaction with internal motion speed. We find no effect of time ( _F_(2,16)=0.011, _p_=.957, $\eta^2$<.001), no effect of internal motion speed ( _F_(2,16)=0.127, _p_=.814, $\eta^2$=.0015) and no interaction ( _F_(4,32)=0.361, _p_=.635, $\eta^2$=.002). This means that we can't replicate the pattern found with the PSE approach. There may be several reasons for this. First, we notice that the trajectory data is highly dependent on participants. It could be that this method provides very noisy estimates, hiding any effect. It could also be that the effects truly vary a lot between participants. Since we can see that some participants do seem to follow the pattern found with the PSE approach, and some show a pattern of increasing average deviation between perceived and true position, it may that the difference between the data sets can simply be explained by random sampling of relatively small numbers of participants.

@@

Some of our participants seem to perceive "jump resets" though, and the model as published, does not predict any such resets. This could be added by including a causal inference step [e.g. @Wei2011]. For the subset of participants that experience jump resets, there may be a process that compares the two position signals (perceived and retinal) and when these exceed some threshold, decides that the two position signals do not share a common source. At that point the perceived position may be discarded, or reset to the retinal position, and the build-up of the illusion may start again. For the remaining participants, either the discrepancy between retinal and perceived location never exceeds their threshold, or there may be another reason they do not employ any causal inference process.

@@

We find that online, manual tracking of the double-drift stimulus is sensitive to illusion strength, but doesn't display clear resets. We do find spontaneous resets in delayed, single-pass re-tracing. The illusion appears to reset, or max out, predominantly at a spatial distance from the true stimulus location, suggesting that resets resolve a source attribution conflict between the true (retinal) and illusory (perceived) position of the stimulus. This confirms the notion that we gauge the position of objects in the world by combining both retinal position signals as well as accumulated position shifts derived from motion signals.
